### 2014-08-29 paper/paper.Rmd
first copy/paste of text. more to come

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.

@@

world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
```

```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
colnames(student2012)[1] <- "name"
# Now add data
student2012$name <- as.character(student2012$name)
student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
student2012$name[student2012$name=="Korea"] <- "South Korea"
student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
student2012$name[student2012$name=="Russian Federation"] <- "Russia"
student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
student2012$name[student2012$name=="China-Shanghai"] <- "China"
student2012$name[student2012$name=="China-Macau"] <- "China"
student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
student2012$PV1MATH <- as.numeric(student2012$PV1MATH)
student2012$PV1READ <- as.numeric(student2012$PV1READ)
student2012$PV1SCIE <- as.numeric(student2012$PV1SCIE)
student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
```

### Teachers, computers, and libraries at school matter

@@

```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Score", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
```

@@

>>>>

This paper will evaluate the role of several factors in which we see inequalities in mathematics, reading, and science outcomes. In section XXXgender, we will evaluate in what subjects we observe gender inequalities. In section XXX we will evaluate inequalities based on the household structure and occupation of the parents. In section XXXpossessions we will look at what home possessions are good predictors of inequalities. Lastly, in sections XXXschool, we will look at the role of school influences.

@@

world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
```
### Teachers, computers, and libraries at school matter

@@

```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
colnames(student2012.sub)[1] <- "name"
# Now add data
student2012.sub$name <- as.character(student2012.sub$name)
student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Score", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
```

@@

It seems that television is the great equalizer. As part of our investigation, we set out to explore how household factors, as indicators of inequality, affect student test scores.

Math scores correlate with the number of household TVs, but how depends on where you live. Nevertheless, access to a quiet study space is generally a plus.

For each country, students’ math scores are compared with number of TVs at home. Countries are grouped by direction and strength of the correlation; groups of countries are mapped (above) and identified (right).

Students are considered by group, and categorized by household factors: number of TVs, and access to a quiet study space.  At far left, number of students in each category; at near left, math scores are summarized for students in each category.

Find your own correlations: http://bit.ly/pisa2012_explorer
